### 2009-04-18 response-to-reviews.txt

> ------------------------------------------------------------------------
> 
> Reviewer #1 (Remarks for the Author):

Both reviews were extremely constructive and a genuine pleasure to
read.

> There are very interesting and potentially quite influential ideas in
> this paper, but too many of them at once are breathlessly muddled up
> together in a fog of words, without sufficiently careful description
> and logical analysis.

We are excited about the possibilities of this project, which we began
3 years ago with the goal of building algorithms to reconstruct
ancient ribosomal RNA.

The main products of this work, and the points of the paper, are:

 1. the mathematics of the new algorithms (tree transducer composition
 and "loopy DP");

 2. implementations of these algorithms as working code (the
 "indiegram" program).

The investigations of various other aspects of RNA reconstruction
(such as error rates, alignment quality & benchmarks of our other
software programs) are secondary to these main algorithmic points.

We agree that the inclusion of these secondary aspects in the main
paper was distracting. Accordingly we have moved these sections to the
supplementary information and trimmed them substantially.

(See below for an explanation of why they were included in the first
place, and what their relevance is; essentially, they are precursors to
future benchmarks we anticipate conducting pending implementation of
more sophisticated MCMC kernel algorithms.)

> The paper does not convey a clear, thorough, reproducible piece of
> science to the reader. It lacks focus, discipline, and rigor. The
> authors need to choose one story they want to tell, then tell it
> completely and thoroughly.

The story we are trying to tell is the systematic extension of
Pair-SCFG-like models to multiple sequences, with a working (albeit
memory-hungry) implementation on a three-branch tree.

We have attempted to bring this story into much sharper focus with our
revised Introduction and structure.

> The paper opens with some grand statements about a research program to
> develop "paleogenetic" algorithms to reconstruct the RNA world. I
> think the authors are using this story to dress up what they're doing,
> rather than really taking this research program seriously and
> analyzing it logically. For starters, in order to phylogenetically
> reconstruct a "primordial ribozyme" such as the "all-RNA primordial
> ribosome" from the "RNA world" from extant sequences, the authors need
> to be able to argue that the last common ancestor of rRNAs lived in
> the RNA world, not the modern DNA/RNA/protein world. It seems clear,
> however, that the last universal common ancestor (LUCA) of all known
> extant life had a reasonably complicated and more or less standard
> DNA/RNA/protein biochemistry, because (even taking abundant lateral
> transfer into account) of how many proteins and biochemical pathways
> can be traced back to LUCA. A serious attempt to reconstruct any
> aspect of the RNA world by phylogenetic analysis must deal with the
> probability that LUCA is almost certainly a bottleneck that occurs
> after the RNA world had passed (if the RNA world existed at all). This
> is not to say that ancestral reconstruction of structural RNAs isn't
> interesting, even if they reconstruct RNAs that lived in less fabulous
> creatures than ribosaurs -- it's only to say that the authors' stated
> goals seem under-analyzed and over-grandiose, and therefore are
> offputting to the reader. Much of the introduction's verbiage about
> the RNA world could be deleted, and the aims of the paper could be
> focused more realistically, concisely, and clearly.

Point taken.

Reconstructing early ribosomes has been a goal of this work for
several years, and Crick's all-RNA ribosome is often cited as a
convincing argument (or at least an inspiration) for the RNA world
hypothesis. So we did not think that, by quoting the RNA world
literature, we were being grandiose or trying to "dress up" our work.

However, we see the reviewer's point that "LUCA != RNA world" and we
agree it would be clearer and more consistent to focus on LUCA.  (Our
algorithms can only reconstruct RNA sequence from periods during which
a "more or less standard biochemistry" was established, and the
reviewers' points about bottlenecks at LUCA are good ones.)

Accordingly, we have rewritten the introduction with more realistic
expectations for the epochs we may actually be able to reach (in fact,
only the first paragraphs of the Introduction and Abstract ever
mentioned the RNA World; the newly revised manuscript barely mentions
it).

We enjoyed the word "ribosaur" -- is it published, or otherwise in
current usage? (Not according to Google...)

We think that the word "paleogenetics" is reasonably well-established
(by the work of, e.g. Pauling & Zuckerkandl, Svante Paabo, Steve
Benner, etc., which we think this naturally relates to); however, if
the reviewer feels that it is confusing or inappropriate, then we can
change the wording accordingly.

> At the same time, the introduction lacks a clear problem definition
> statement. What is the input; what is the desired output; what is the
> main idea behind the approach for transforming input to output?

The central problem, as we now state more clearly in the Introduction,
is this:

 Given a conditionally normalized Pair SCFG, modeling P(Y|X) where X
 and Y are foldback RNA structures, along with a phylogenetic tree,
 what is the corresponding multiple-sequence SCFG; and can we run the
 Cocke-Younger-Kasami algorithm on it (suitably constrained) ?

The solution to this problem is given by the transducer composition
algorithm, which is the core result of this paper.

> Throughout the paper, terms and programs are thrown around in a
> muddled soup, as if everything is a general purpose magic inference
> tool. _xrate_ estimates substitution rate parameters (p.2) and folds
> RNA (p.5); _evoldoer_ generates pair SCFGs (p.3) and "implements"
> TKFST (p.5), whatever "implements" may mean (any given probabilistic
> model may be used for a wide range of different simulations or
> inference problems). Do they slice and dice tomatoes too? At far too
> many points, the authors recite lists of possible applications that
> might be tackled someday, or lists of other approaches that can be
> viewed as special cases of their approach -- without ever particularly
> clearly defining what their approach is. I'm all in favor of
> generalizable approaches and powerful probabilistic models, but I want
> to see at least one specific, well-described problem solved before one
> expounds on all the other things that might possibly be done. That's
> what the Discussion is for.

We have emphasized the central problem of constructing a model of
multiple sequences given a model of the evolution of two sequences and
a phylogenetic tree (the "transducer composition" algorithm) in the
revised manuscript, including a full statement of inputs and outputs.

We have also removed excessive references to program names as
requested by both reviewers.

We have attempted to clarify the possible applications of these
methodologies and place them in the context of prior work.  Some of
these applications/questions are:

-- Given a conditional distribution P(Y | X), can one invert this and
compute (or sample from) the distribution of the ancestral sequence,
i.e. P(X | Y)?

-- Can one repeatedly apply this model and e.g. sample Z from P(Z|Y)
where Z is also a sequence/structure? Can one sum over all values of Y
and compute P(Z|X) directly, or sample Z from it?  (This is indirectly
related to the problem of time-dependence, since the branch from X-->Z
represents twice as much evolutionary time as the branch from X-->Y.)

-- How well does the model handle multiple sequences and missing data?
E.g. can we sample from the distribution of unobserved "grandparent"
sequences, given a set of observed "grandchildren" (and unobserved
"parents")?

-- What about if you have an unobserved parent with three observed
children? Can you sample from P(parent|children) ?  (This is closest
to the task we actually implement; doing this is sufficient to address
all of the other questions using an MCMC approach, as shown by our
2001 paper and several other researchers.)

-- Can we estimate the multiple alignment and phylogeny of a given
set of sequences, starting only with the model P(Y | X) ?

All of these questions can be addressed by the multi-sequence SCFGs we
describe, using standard SCFG algorithms (Inside-Outside, CYK, etc.)
and constrained MCMC variants on those algorithms.

> The authors could do a much better job of recognizing that ancestral
> sequence reconstruction has been done for decades: both parsimony
> methods and maximum likelihood methods for phylogenetic inference
> explicitly infer sequences at ancestral nodes in a tree.

Thank you for pointing this out; we did not intend to slight previous
authors.  In our rewritten Discussion, we have cited more of the
basic papers that established the computational and mathematical
foundations of the field.  We also cite the 2007 book edited by David
Liberles, an excellent survey of current uses of substitution models
for ancestral reconstruction, as well as the origins of the field
(e.g. Pauling & Zuckerkandl's paper, parsimony & ML methods, and so on).

> What's of interest here is whether that can be done not only for
> residue substitution events, but for insertion/deletion events as
> well,

We completely agree.
Insertions and deletions in unstructured sequences can be well-handled by
conditionally-normalized Pair HMMs, or "string transducer" state machines,
for which several implementations of the transducer composition
algorithm for building multi-sequence models from pairwise models now exist.
One of the central goals of this paper was to extend this theory to
the case of RNA secondary structures (or parse-tree transducers), in
order to model effects like indels that conserve RNA structure.

> and whether the inferences are sufficiently accurate that we can
> take them seriously and actually synthesize the inferred ancestor(s)
> and believe that they tell us something about the structure and
> function of extinct biomolecules.
> 
> The Methods are unnecessarily opaque and confusing, focusing too much
> on peripheral terminology and too little on an actual problem at
> hand. For example, for all this talk of "transducer theory", it is
> never clear how a "transducer" is meaningfully different from any
> other conditional probability model used in sequence analysis -- such
> as profile HMMs and profile SCFGs, which model a conditional
> distribution P(Y | X) via a model built from sequence X, P(Y | X) ~=
> P(Y | model) P(model | X) using a point estimate of an appropriate
> model derived from X, where X could be either a query
> sequence/alignment in a similarity search,

We apologize for the lack of clarity.  We have attempted to rectify
this in the rewritten manuscript by first discussing
jointly-normalized models (Pair HMMs) and then introducing
conditionally-normalized models (transducers).

We believe that the transducer is a meaningful improvement on
HMMs/SCFGs because of the transducer composition algorithm, which
offers a natural way to construct models of multiple sequences; for
example, given a model P(Y|X), and a suitable phylogeny, we can build
a model like this

   P(A, B, C, D, E | X, TREE)

where X is the root sequence of the tree, and {A,B,C,D,E} are leaves.
This model correctly accounts for all combinations of overlapping
indel events that can possibly occur on the branches of the tree --
for example, an insertion on an ancient branch can be interrupted by
an insertion on an intermediate branch, then both insertions can be
[partially] deleted on an even younger branch. The model is a
multi-sequence HMM, so that multiple sequence alignment is simply an
application of the Viterbi algorithm (or posterior decoding), and MCMC
Gibbs samplers can be constructed by applying suitable constraints to
the Forward algorithm.

Furthermore, dynamic programming (and MCMC-sampling) algorithms exist
to invert this model

   P(X | A, B, C, D, E, TREE)

Ancestral reconstruction involves sampling from and/or maximizing this
inverted distribution. Many other Bayesian manipulations can similarly
be performed on this likelihood function.

This ability to generalize systematically to multiple sequences is
missing from HMMs and SCFGs. While it is possible to convert a
single-sequence HMM into a phylo-HMM (by having it emit alignment
columns instead of single symbols), and even to simulate gaps to some
extent (by modeling them as an additional character), there is no
simple recipe (other than transducer composition) for extending a Pair
HMM to a corresponding multi-sequence HMM that includes affine gap
penalties, overlapping indels, nested indels, etc.

> model derived from X, where X could be [...]
> an ancestor on a tree in a phylogenetic inference problem.

We appreciate the point that there are similarities between Profile
SCFGs and the transducers which we describe here. For example, the
paper by Klein & Eddy (2003), describing the RSEARCH program,
introduces a profile SCFG which may be regarded as modeling a
conditional distribution P(Y|X). However, to our knowledge, existing
literature on Profile SCFGs does not describe the formalisms which we
need to reconstruct structured sequences on trees.  In particular, we
need ways to find:

-- the corresponding joint Pair-SCFG, modeling P(X, Y);
-- the Bayesian inverse Profile-SCFG, modeling P(X | Y);
-- situations involving more than two sequences, e.g. the
   Profile-SCFG for P(X | Y, Z), where X is the last common ancestor
   of Y and Z; or the corresponding Pair-SCFG for P(Y, Z); or the
   corresponding Triplet-SCFG for P(X, Y, Z).

Transducers provide a generic, systematic way to build such
multi-sequence, phylogenetically-structured SCFGs. In this paper we
attempt to describe how to perform these procedures.

> Seems to me that the critical thing,
> which the authors actually _underemphasize_, is how one makes this
> conditional probability model P(Y | X) be dependent on evolutionary
> time t, P(Y | X,t), via an underlying continuous-time Markov model of
> evolution in terms of rates of individual events including
> substitutions, insertions, and deletions.

This is a good point.  Time-dependent parameterization is indeed
critical, but it can be handled as a separate problem from the one we
address here.  Our goal is to provide a modular solution to the
ancestral reconstruction problem, wherein the problem of creating
accurate (two-sequence) models can be treated as disjoint from the
problem of extending those models to multiple sequences.  In this
paper we have focused on the second problem: Given a model of the
evolution of two sequences, how do we construct a model of many
sequences?

For this reason, while the time parametrization problem is certainly
central to the ancestral reconstruction problem, we believe that it
can be treated as a distinct problem from the multi-sequence model
construction problem which is the focus of this paper.  As the
reviewer points out, we largely gloss over how to choose an
appropriate time parametrization, and instead focus on:
Suppose you do have such a time-dependent model. How do you then
extend it from two sequences to multiple sequences?

More generally, how do you sample from some generic distribution

P(X_1, X_2 ... X_M  |  Y_1, Y_2 ... Y_M,   TREE)

where {X_1 ... X_M} are all ancestors, and {Y_1 ... Y_M} are all
descendants, related by a given TREE?

For substitution models, the answer to this question involves
Felsenstein's pruning algorithm.

For indel/RNA models, our answer is that you construct a
multi-sequence SCFG and then sample paths through it (appropriately
conditioned on the data {Y_n}).  Transducers allow you to build this
Multi-SCFG, assuming you already know the Pair-SCFG (i.e. the
parse-tree transducer).

Note that neither of these answers is the same as the problem of
getting time-dependent parameters! For substitution processes, you get
time-dependent parameters by matrix exponentiation (i.e. solving a
continuous-time Markov chain).

For RNA/indel processes, the TKFST model described in earlier work
(Holmes 2004) represents one attempt to get to these time-dependent
parameters for point substitutions, covariant substitutions,
insertions, deletions and structural changes.

TKFST is flawed, and certainly needs work to make it more realistic;
however, we believe that it's a reasonable simple model to test our
solution to the second half of the reconstruction problem.

TKFST is not the main focus of the present paper -- it's just the
particular pairwise model we're using as an example.  We could have
used another pairwise model to illustrate transducer composition, and
we have attempted to make sure that the composition algorithm itself
is not presented in a TKFST-specific way.

> This time-dependent evolutionary model has essentially nothing to do
> with (is separable from) "transducer theory" as far as I understand;

We agree; with the exception that, if you take the time-dependent model as
representing a single branch, and consider extending it to an
arbitrary phylogenetic tree using the sum-product algorithm, then this
line of reasoning will lead you to an algorithm which is essentially
the same as the Inside-Outside algorithm on a Multi-SCFG constructed
by transducer composition.

For example, the multidimensional DP algorithm of Hein (PSB, 2001) for
computing the TKF model likelihood on a binary phylogenetic tree is
exactly equivalent to the a Forward algorithm for a composite
phylo-transducer. Though not originally expressed in transducer
formalism, Hein's algorithm is (arguably) easier to explain that way.

So there are conceptual relationships. It's also true that serial
composition of transducers amounts to "adding" two branch lengths, so
this is another relationship to the time-dependent theory (indeed, it
can be considered a "test" of the additive property of time-dependent
models).

But these are subtle connections. We agree with the reviewer that these
two questions (branch length and tree topology) can for most intents &
purposes be considered to be separable -- just as, with substitution
models, matrix exponentiation is separable from pruning.

> "transducer theory" is just a different set of words for conditional
> probability models, with the evolutionary continuous-time Markov
> model providing the particular _parameters_ that get plugged into
> the conditional probability model.
> (That is, I believe that I can interchange a
> "string transducer" with an HMM and "parse tree transducer" with an
> SCFG with no loss of information. If this is true, the exact names of
> things aren't important so long as they're clearly defined and
> explained for the particular purpose at hand, and one shouldn't be
> trying so hard to brand "transducer theory" as new. If I'm wrong, I
> would like to see the authors' clear explanation.)

While they are very similar, the essential difference is that there is
no general-purpose recipe to convert a Pair-HMM or Pair-SCFG into a
multiple-sequence HMM or SCFG. In order to do this you would need

(a) to label one sequence as the ancestor and the other as the
descendant;
(b) to ensure that the HMM/SCFG was then conditionally normalized;
(c) to devise an appropriate algorithm for constructing the state
space and the rule space of the composite machine (which essentially
amounts to co-ordinating transitions between the component pairwise
machines).

Steps (a) and (b) are trivial, but step (c) is nontrivial and is the
point of this paper. (The analogous algorithm for string transducers
is described in (Holmes, 2003).)

The reviewer is correct that, e.g., Pair HMMs and "string transducers"
are very similar, so that a conditionally-normalized Pair HMM is
essentially the same as a string transducer. This just amounts to a
restatement of steps (a) and (b); the innovation is in step (c).

We did not aim arbitrarily to brand transducers as new; rather, we are
attempting to highlight connections to existing literature.  We have
attempted to make this clearer in the revised manuscript by
emphasizing the term's origin in the computational linguistics
literature ("parse-tree transducer" comes from the same literature).
We have also attempted to make clearer the tight connections between
existing HMM theory in bioinformatics and the "transducer theory"
described in computational linguistics (c.f. Mehryar Mohri et al.)

The idea of transducer composition, above and beyond the
straightforward idea of a conditional probability model, has been used
in several other recent papers. For example:

(1) Paten B, Herrero J, Fitzgerald S, Beal K, Flicek P, Holmes I,
Birney E.  Genome-wide nucleotide-level mammalian ancestor
reconstruction.  Genome Res. 2008.

(2) Satija R, Pachter L, Hein J.  Combining statistical alignment and
phylogenetic footprinting to detect regulatory elements.
Bioinformatics. 2008 May 15;24(10):1236-42.

(3) Varadarajan A, Bradley RK, Holmes IH.  Tools for simulating
evolution of aligned genomic regions with integrated parameter
estimation.  Genome Biol. 2008 9(10):R147.

We think that the transducer formalism makes these papers considerably
simpler to read (and provides a connecting thread between them).

### 2009-04-19 paper.tex
fixed typo in null cycle elimination section

Let $\leftside{A}$ be the set of rules that can be applied to nonterminal $A \in \gnons$ (i.e., rules in which $A$ appears on the left),
and let $\rightside{B}$ be the set of rules that can generate nonterminal $A$ (i.e.,rules in which $A$ appears on the right).
Let $\allrules{B} = \leftside{A} \cup \rightside{A}$ be the set of all rules involving $A$.
Define these also on sets of nonterminals, e.g., $\leftside{\gnulls} = \bigcup_{A \in \gnulls} \leftside{A}$ for $\gnulls \subseteq \gnons$.

>>>>

Let $\leftside{A}$ be the set of rules that can be applied to nonterminal $A \in \gnons$ (i.e., rules in which $A$ appears on the left),
and let $\rightside{A}$ be the set of rules that can generate nonterminal $A$ (i.e.,rules in which $A$ appears on the right, including bifurcations which also generate another nonterminal as well as $A$).
Let $\allrules{A} = \leftside{A} \cup \rightside{A}$ be the set of all rules involving $A$.
Define these also on sets of nonterminals, e.g., $\leftside{\gnulls} = \bigcup_{A \in \gnulls} \leftside{A}$ for $\gnulls \subseteq \gnons$.
